Lazy Linearity for a Core Functional Language

### 2025-10-15 popl26.tex
Up to start of section 2

Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a linearly typed Core would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.
%
% Linear core actually not so good
% Additionally, while not yet a reality, linearity in Core could be used to inform
% certain program optimisations, i.e. having linear types in Core could be used to
% further optimise certain programs and, therefore, benefit the runtime
% performance characteristics of our programs. For example, Linear Haskell\cite{}
% describes as future work an improvement to the inlining optimisation: Inlining
% is a centerpiece program optimisation primarily because of the subsequent
% optimising opportunities unlocked by inlining. However, it relies on a heuristic
% process known as \emph{cardinality analysis} to discover safe inlining
% opportunities. Unfortunately, heuristics can be volatile and fail in identifying
% crucial inlining opportunities resulting in programs that fall short of their
% performance potential. On the contrary, the linearity information could be
% integrated in the analysis and used as a (non-heuristic) cardinality
% \emph{declaration}.
%
% While the current version of Core is linearity-aware (i.e.,~Core has so-called
% multiplicity annotations in variable binders), its type system does not (fully)
% validate the linearity constraints in the desugared program and essentially
% fails to type-check programs resulting from several optimising transformations
% that are necessary to produce efficient object code. The reason for this latter
% point is not evidently clear:
% %
% if we can typecheck linearity in the surface level Haskell why do we fail to do
% so in Core?
%
%
However, despite the advantages of having linear types in Core, the status quo is
that linearity is effectively ignored in Core, being only checked
in the source Haskell code. The reason is that, after desugaring, the
Core-to-Core optimising transformations eliminate and rearrange most of the
syntactic constructs through which linearity checking is performed -- often
resulting in programs that are syntactically very different from the original,
especially due to the flexibility provided by laziness with regard to the
optimisations a compiler may perform.

While compiler optimisations aim to preserve the semantics of
programs and therefore a program's resource consumption,
a \emph{naive} syntactic analysis of the optimised programs (as
performed by the current Core type checker) often fails to recognize
the program as linear.
%
As a trivial example, let $x$ be a linear resource in the next two expressions,
where the latter results from inlining $y$ without eliminating the
binder (a common optimising transformation in GHC). Even
though the second expression no longer appears linear (as there are now two occurrences
of $x$), it is indeed linear because the let-bound expression is never
evaluated under \emph{lazy evaluation} and $x$ is still consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and design a linear type
system for Core which accepts the vast majority of programs and optimising
transformations that, while correct under non-strict evaluation, are rejected
by traditional linear typing disciplines.
%
%
Additionally, we show \ROUNDTWO{that} our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

\item We introduce the concept of \emph{semantic} linearity, in
contrast to \emph{syntactic} linearity in Haskell, by example
(Section~\ref{sec:linearity-semantically}).
%
A precise definition of \emph{semantic} linearity is given by our instrumented
operational semantics, under which a program using a linear resource
non-linearly gets stuck (Section~\ref{sec:main:metatheory}).

\item We introduce Linear Core, a non-strict linear language with
the key features of Core (except for type equality coercions), whose type system
accounts for semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to capture linearity
semantically in this context (Section~\ref{sec:main:linear-core}).

@@

\noindent Additional proofs and definitions are given in
Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.

@@

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linearity, Semantically
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linearity, Semantically\label{sec:linearity-semantically}}

A linear type system typically guarantees that linear resources are
consumed exactly once. Consequently, whether a program is
well-typed under this lens intrinsically depends on the
definition of consuming a resource.
In general, consuming a resource is equated with its 
\emph{syntactic occurrence} along a control-flow path in the program. However, as this section
makes clear, in linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following Haskell-like program defining a linear
function \lstinline{f} (i.e., a function that should use its argument
exactly once), there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other is returned. If the program is evaluated lazily,
the handle will not be closed because the computation that closes the handle is
never evaluated (the non-linear \lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

@@

\subsection{Semantic Linearity by Example\label{sec:semantic-linearity-examples}}

>>>>

Just as Core's type system \cite{10.1145/1190315.1190324} provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a linearly typed Core would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.
%
% Linear core actually not so good
% Additionally, while not yet a reality, linearity in Core could be used to inform
% certain program optimisations, i.e. having linear types in Core could be used to
% further optimise certain programs and, therefore, benefit the runtime
% performance characteristics of our programs. For example, Linear Haskell\cite{}
% describes as future work an improvement to the inlining optimisation: Inlining
% is a centerpiece program optimisation primarily because of the subsequent
% optimising opportunities unlocked by inlining. However, it relies on a heuristic
% process known as \emph{cardinality analysis} to discover safe inlining
% opportunities. Unfortunately, heuristics can be volatile and fail in identifying
% crucial inlining opportunities resulting in programs that fall short of their
% performance potential. On the contrary, the linearity information could be
% integrated in the analysis and used as a (non-heuristic) cardinality
% \emph{declaration}.
%
% While the current version of Core is linearity-aware (i.e.,~Core has so-called
% multiplicity annotations in variable binders), its type system does not (fully)
% validate the linearity constraints in the desugared program and essentially
% fails to type-check programs resulting from several optimising transformations
% that are necessary to produce efficient object code. The reason for this latter
% point is not evidently clear:
% %
% if we can typecheck linearity in the surface level Haskell why do we fail to do
% so in Core?
%
\ROUNDTWO{However, despite the advantages of having linear types in Core, the status quo is}
that linearity is effectively ignored in Core, being only checked
in the source Haskell code. The reason is that, \ROUNDTWO{after} desugaring, the
Core-to-Core optimising transformations eliminate and rearrange most of the
syntactic constructs through which linearity checking is performed -- often
resulting in programs that are syntactically very different from the original,
especially due to the flexibility provided by laziness with regard to the
optimisations a compiler may perform.

While compiler optimisations aim to preserve the semantics of
programs and therefore a program's resource consumption,
a naive syntactic analysis of the optimised programs as
performed by the current Core type checker often fails to recognize
\ROUNDTWO{resulting} programs as linear.
%
As a trivial example, let $x$ be a linear resource in the next two expressions,
where the latter results from inlining $y$ without eliminating the
binder -- a common optimising transformation in GHC. Even
though the second expression no longer appears linear (as there are now two occurrences
of $x$), it is indeed linear because the let-bound expression is never
evaluated under \ROUNDTWO{non-strict} evaluation and $x$ is still consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
\ROUNDTWO{linear} type system: it does not account for the non-strict evaluation model
of Core and, thus, a whole class of programs that are linear under the lens of
non-strict evaluation are rejected by Core's current \ROUNDTWO{checker}.
%
In this work, we address this limitation by exploring how syntactic linearity
breaks down in the presence of non-strictness and design a linear type
system for Core which accepts the vast majority of programs and optimising
transformations that, while correct under non-strict evaluation, are rejected
by traditional linear typing disciplines.
%
%
Additionally, we show \ROUNDTWO{that} our system
%
% and its implications on validating and influencing optimising transformations
% by designing a type system which analyses how syntactic linearity breaks down
% in the presence of laziness
is suitable for the intermediate language of an
optimising compiler by implementing it as a Glasgow Haskell Compiler plugin.
%
Our contributions are as follows:
%
\begin{itemize}

\item We \ROUNDTWO{illustrate} the \ROUNDTWO{lazy semantics} of linearity, in
contrast to the \ROUNDTWO{traditional} syntactic \ROUNDTWO{understanding
of} linearity, in Haskell, by example (Section~\ref{sec:linearity-semantically}).
%
A precise \ROUNDTWO{semantic definition of} linearity, \ROUNDTWO{closely
following that of Linear Haskell~\cite{cite:linearhaskell}}, is given by
our instrumented operational semantics, under which a program using a
linear resource non-linearly gets stuck
(Section~\ref{sec:main:metatheory}).

\item We introduce Linear Core, a \ROUNDTWO{lazy} linear language with
the key features of Core (except for type equality coercions), whose type
system \ROUNDTWO{accepts the lazy} semantics of linearity. To the best of our
knowledge, this is the first type system \ROUNDTWO{to accept programs
which depend on non-strict evaluation to be understood as linear}
(Section~\ref{sec:main:linear-core}).

@@

\noindent \ROUNDTWO{While GHC's Core language is our main focus, we
note that the ideas explored in this work can potentially be applied to any
language with non-strict features, even if the language itself is strict.}

Additional proofs and definitions are given in \ROUNDTWO{TODO}
Appendices~\ref{app:proofs},~\ref{app:optimisations} and~\ref{app:opsem}.

@@

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% {{{ Linearity, Semantically
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Linearity, Lazily\label{sec:linearity-semantically}}

Linear type systems guarantee that linear resources are consumed exactly once.
\ROUNDTWO{Thus}, whether a program is \ROUNDTWO{accepted as linear} intrinsically depends on
the definition of consuming a resource.
%
\ROUNDTWO{Traditionally}, consuming a resource \ROUNDTWO{has been} equated with its 
syntactic occurrence along a control-flow path in the program. However, as this
section \ROUNDTWO{aims to} make clear, in linear type systems for non-strict languages, a
distinction between using resources \emph{syntactically} and \emph{semantically} is needed.
%
For instance, in the following Haskell-like program defining a linear
function \lstinline{f} (i.e., a function that should use its argument
exactly once), there are two syntactic
occurrences of \lstinline{handle}: one is in a let-bound computation that
closes this handle, the other is returned. If the program is evaluated lazily,
the handle will not be closed because the computation that closes the handle is
never evaluated (the non-linear \lstinline{x} goes unused).
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
% Intuitively, a computation that depends on a linear resource to produce a
% result consumes that resource iff the result is effectively computed; in
% contrast, a computation that depends on a linear resource, but is never run,
% will not consume that resource.
%
% Even though there are two occurrences of \lstinline{handle} in the program, and
% it looks as though the \lstinline{handle} may be closed before returned, this function may
% actually only \emph{consume} used once -- it depends on the evaluation strategy of the
% language -- the \lstinline{handle} is closed iff the let bound computation is
% evaluated.
%
%
% Intuitively, \emph{consuming} a resource is not necessarily synonymous with its
% syntactic occurrence. In the example, \emph{non-strict} evaluation means the
% handle is never closed by the function.
%
%
% Depending on the evaluation strategy of the language, a computation that closes
% a handle might not be evaluated, and thus is not consumed.
% With \emph{eager} evaluation semantics,
% the let
%   bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
% returned. It is clear that a linear type system should not accept such a
% program since the linear resource \lstinline{handle} is duplicated -- it is used in a
% computation that closes it, while still being made available to the caller of
% the function;
% %
% on the other hand, with \emph{call-by-need},
%
% Reconsider the \lstinline{handle} example program under
% \emph{call-by-need} evaluation semantics:
% The let bound expression is only evaluated if the binding \lstinline{x} is
% needed. The function returns the \lstinline{handle} right away, and forgets the
% let bound \lstinline{x} as its scope is exited, such that \lstinline{x} will
% never be needed -- consequently, the \lstinline{handle} is never closed by the
% function...
%
%
%

@@

\subsection{Lazy Linearity by Example\label{sec:semantic-linearity-examples}}
